**Fabryka**

### 2025-04-13 fabryka_pci/article.Rmd
last changes in the rmarkdown article

Marc THOMAS ^1\*^ \bigbreak \bigbreak

@@

E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
\bigbreak \bigbreak

**Key words** : R shiny web application, fabric, site formation
processes, spatial analyses, orientation and inclination \bigbreak
\bigbreak

@@

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using *shiny*
package [@chang2024], and this article, written in rmarkdown
[@allaire2014; @xie2018; @xie2020], are both available online in a
public github repository
(<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing
this article, you can either clone the repository, or push on the
changes via github.\pagebreak

@@

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It
is a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages *e.g.* [@isaac1967; @schick1986;
@thomas2019]. Objects abandoned by prehistoric people can be considered
as sedimentary particles and are affected by the same processes
[@schiffer1983]. As geological particles are often not elongated
enough, the fabric analysis made of the sedimentary deposits depends
mainly on the archaeological material (bones and lithics), which enables
them to be inserted into the formation dynamics of the site and to
obtain information on their representativeness for the reconstruction of
past activities.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the direction of flow, with a slight upstream
inclination. Since G. Isaac's pioneering work, there has been an
abundant literature devoted to the analysis of fabrics *e.g.*
[@benn1994; @lenoble2004; @mcpherron2018] allowing this method to be
increasingly performed on archaeological sites. Several software allow
data processing, some of which are not freely accessible, do not allow
spatial exploration of the data or do not offer interfaces to facilitate
the task of non-programmer users *e.g* [@mcpherron2018].

@@

During excavation, the fabric of elongated remains is generally
systematically measured. The direction (= bearing) of the lowest point
is measured in the range [0; 360] with North as the reference (i.e.
direction = 0°). The inclination (= plunge, dip) is measured in the
range [0; 90] with horizontal as a reference. The direction [0; 360] can
be converted into an orientation measured in the range [0; 180] by
subtracting 180° from the measurements in the range ]180; 360].

@@

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>). The
device stores and associates data with the order of measurement. When
the device is tilted downwards, the direction of the lowest point is
measured and, the recorded inclination is negative. When the device is
tilted upwards, the direction of the highest point is recorded, and the
inclination is positive. To obtain data that is suitable for statistical
analysis, in the first case the absolute value of the inclination must
be retained. In the second case (*i.e.* positive inclination), if the
measured direction is in the interval [0; 180], 180° must be added to
the direction, and if the measured direction is in the interval ]180;
360], 180° must be subtracted from the direction.

@@

1.  The first considers the orientation only. The calculation of the
    orientation rate "r" (or "L" when expressed as a percentage, Curray
    formula [@curray1956] tests the hypothesis of a unimodal
    distribution of orientations. If the "p" value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the orientation rate "r" where '$\alpha$' is
    the orientation in radians and "n" is the sample size is as follows:

@@

The method consisting of doubling the value of the orientations before
calculating the orientation rate L and the Rayleigh test, as proposed by
Krumbein [@krumbein1939], tests the hypothesis of a bimodal
distribution of orientations with a period of 90°. The hypothesis of a
90° period bimodal distribution of a sample of orientations is retained
when p \< 0.05 [@curray1956]. The sensitivity of the Rayleigh test to
the sample size 'n' prevents samples from being compared with each other
when 'n' varies [@curray1956; @bertran2002]. For example, when the
orientation rate "L" is 25%, the Rayleigh test gives p-values below a
critical threshold of 0.05 for series with more than 50 measurements
(fig. \ref{fig:rayleigh}).

@@

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n'", fig.height = 3.5, fig.pos="f"}

@@

To distinguish different statistical populations, a spatial method
proposed by McPherron (2018) is used. Thanks to this methods it is
possible to calculate the Benn indices and to test the orientation rate
'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
remains, and its n nearest neighbours in 3 dimensions (R McPherron 2018
script, modified , formula (**4.**)). The analysis of the fabric as
proposed by McPherron (2018) has the advantage of spatialising the
measurements and thus avoiding the arbitrary sampling classically
carried out by square or by décapage. It is possible to localize the
statistical signal of fabrics by varying the number of measurements
sampled to calculate the Benn indices. The choice of about 40 to 60
measurements (each remain and its 40 to 60 nearest neighbours) to
calculate the Benn indices allows local particularities to be observed,
and above all, it allows the Rayleigh test to be carried out in its
field of application (stability of the test between around 40 to 50
measurements after [@bertran2002]).

@@

As proposed by McPherron (2018), a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections of the material along the
axes of the excavation grid. Projections (xy, xz and yz axes) are
performed. \bigbreak

@@

## Upload your data

@@

Five cases (i.e. forms of data) are covered by the application (fig.
\ref{fig:figure_upload_your_data}D):

@@

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

If you miss a context column, please create it by repeating a same
character.

@@

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters (samples
and code) and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see [@thomas2019] and [@texier2020]). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. In the rose
diagram, orientations are considered (and not directions) as the
Rayleigh test is perform on orientations (see Curray 1956) in the
summary table. You can download all the figures by selecting a file
format (.pdf, .jpeg and .png) and pressing the 'download' button (fig.
\ref{fig:figure_classical_method}E and F).

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

@@

-   Benn diagrams with Benn indices computed with nearest neighbours
    points (McPherron 2018)
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours = one line) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

@@

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

@@

\begin{figure} [H]
\centering
  \includegraphics[width=17cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}

\pagebreak

@@

The 'spatialised' method generates often more linear fabrics in
comparison with the 'classical' method. This can be explained by the
fact that objects that are close in space are more likely to resemble
each other than objects that are further in space. \bigbreak

@@

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new "spatialised" reference
model (*cf.* the "classical" model proposed by Bertran and Lenoble 2002).
Some processes are already available. \bigbreak

@@

The contributors panel aims at citing all contributors of this project
which is open to everyone interested in the fabric analysis methods or
to contribute in building a new spatial model by sending reference data.
If any request or contribution, please contact the author of this
article. \bigbreak

@@

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. It also benefited
from the help of Rémi Lemoy and François and Victor Baleux on
mathematical issues. A first version of the application was tested by
Emmanuel Discamps and Aurélien Royer. I thanks them for their advice and
support. \bigbreak

-   use Fabryka $\beta$ 3.0: <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source codes: <https://github.com/marchaeologist/fabryka/tree/main> \smallbreak
-   after the review process, create a doi (https://zenodo.org/login/?next=%2Faccount%2Fsettings%2Fgithub%2F) for each app' version. 
    \pagebreak

>>>>

\begin{figure} [H]
\centering
  \includegraphics[width=4cm]{figure/hex_Fabryka4.png}
\label{}
\end{figure}

Marc THOMAS ^1\*^ (https://orcid.org/0000-0002-8160-1910) \smallbreak

@@

E-mail: [marcthomas1\@hotmail.fr](mailto:marcthomas1@hotmail.fr){.email}
\bigbreak

**Key words** : R shiny web application, fabric, site formation
processes, spatial analyses, orientation and inclination \bigbreak

@@

You can use Fabryka $\beta$ 3.0 by following this link
<https://marchaeologist.shinyapps.io/fabryka/>. The version will be
updated to Fabryka 1.0 after the review process of this article. The
Fabryka application files, made in R [@rcoreteam2025] by using *shiny* package [@chang2024], and this article, written in rmarkdown [@allaire2014; @xie2018; @xie2020], are both available online in a public github repository (<https://github.com/marchaeologist/fabryka/tree/main>). When reviewing this article you can fork and clone the fabryka project repository on github to make suggestions for modifications.\pagebreak

@@

Fabric is made of the orientation and inclination of elongated
sedimentary particles [@bertran1997; @bertran2002; @lenoble2004]. It is
a statistic (*i.e.*, it needs as much measurements as possible) that
enables the identification of sediment sources that make up the deposits
[@texier2020], sedimentary processes involved in their formation
[@bertran1997; @bertran2002; @lenoble2004] and their impact on
archaeological assemblages [@isaac1967; @schick1986; @thomas2019].
Objects abandoned by prehistoric people can be considered as sedimentary
particles and are affected by the same processes [@schiffer1983]. As
geological particles are often not elongated enough, the fabric analysis
made of the sedimentary deposits depends mainly on the archaeological
material (bones and lithics), which enables them to be inserted into the
formation dynamics of the site and to obtain information on their
representativeness for the reconstruction of past activities.

One of the first archaeological application probably dates to the
experiments carried out by the South African G. Isaac (1967) on the
shores of Lake Magadi in Kenya. G. Isaac wanted to explain the role
played by hydrological processes on the bone accumulations found in the
Great Rift Valley. In November 1964, he placed several concentrations of
experimental material near a river branch. A few months later, heavy
rains had reactivated the valley's rivers. In addition to the
displacement or even loss of all or part of the material, G. Isaac
revealed a reorientation that was mainly perpendicular and, to a lesser
extent, parallel to the direction of flow, with a slight upstream
inclination. Since G. Isaac's pioneering work, there has been an
abundant literature devoted to the analysis of fabrics [@benn1994;
@lenoble2004; @mcpherron2018] allowing this method to be increasingly
performed on archaeological sites. Several software allow data
processing, some of which are not freely accessible, do not allow
spatial exploration of the data or do not offer interfaces to facilitate
the task of non-programmer users [@mcpherron2018].

@@

During excavation, the fabric of elongated remains is generally
systematically measured. The direction (= bearing) of the lowest point
is measured in the range [0; 360] with North as the reference (*i.e.*
direction = 0°). The inclination (= plunge, dip) is measured in the
range [0; 90] with horizontal as a reference. The direction [0; 360] can
be converted into an orientation measured in the range [0; 180] by
subtracting 180° from the measurements in the range ]180; 360].

@@

Recently, speleologist have created a digital inclinometer (Leica
DistoX310 modified to DistoX2; <https://paperless.bheeb.ch/>) which can
be used to measure fabrics [@discamps2024]. The device stores and
associates data with the order of measurement. When the device is tilted
downwards, the direction of the lowest point is measured and, the
recorded inclination is negative. When the device is tilted upwards, the
direction of the highest point is recorded, and the inclination is
positive. To obtain data that is suitable for statistical analysis, in
the first case the absolute value of the inclination must be retained.
In the second case (*i.e.* positive inclination), if the measured
direction is in the interval [0; 180], 180° must be added to the
direction, and if the measured direction is in the interval ]180; 360],
180° must be subtracted from the direction.

@@

1.  The first considers the orientation only. The calculation of the
    orientation rate 'r' (or 'L' when expressed as a percentage,
    Curray's [-@curray1956] formula tests the hypothesis of a unimodal
    distribution of orientations. If the 'p' value returned by the
    Rayleigh test is less than a threshold of 0.05, the null hypothesis
    of a random orientation from a uniform distribution is rejected. The
    formula for calculating the orientation rate 'r' where '$\alpha$' is
    the orientation in radians and 'n' is the sample size is as follows:

@@

The method consisting of doubling the value of the orientations before
calculating the orientation rate L and the Rayleigh test, as proposed by
Krumbein [-@krumbein1939], tests the hypothesis of a bimodal distribution
of orientations with a period of 90°. The hypothesis of a 90° period
bimodal distribution of a sample of orientations is retained when p \<
0.05 [@curray1956]. The sensitivity of the Rayleigh test to the sample
size 'n' prevents samples from being compared with each other when 'n'
varies [@curray1956; @bertran2002]. For example, when the orientation
rate "L" is 25%, the Rayleigh test gives p-values below a critical
threshold of 0.05 for series with more than 50 measurements (fig. 2).

@@

```{r rayleigh, echo = FALSE, warning = FALSE, message = FALSE, fig.cap="Evolution of the p-value returned by the Rayleigh test as a function of the orientation rate 'L' and the sample size 'n' \\label{rayleigh}", fig.height = 3.5, fig.pos="f"}

@@

To distinguish different statistical populations, a spatial method
proposed by McPherron (2018) is used. Thanks to this methods it is
possible to calculate the Benn indices and to test the orientation rate
'L' (formula (**3.**) modified R [@mcpherron2018] script) for each
remains, and its n nearest neighbours in 3 dimensions (McPherron
[-@mcpherron2018] R script, modified , formula (**4.**)). The analysis
of the fabric as proposed by McPherron [-@mcpherron2018] has the
advantage of spatialising the measurements and thus avoiding the
arbitrary sampling classically carried out by square and décapage. It
is possible to localize the statistical signal of fabrics by varying the
number of measurements sampled to calculate the Benn indices. The choice
of about 40 to 60 measurements (each remain and its 40 to 60 nearest
neighbours) to calculate the Benn indices allows local particularities
to be observed, and above all, it allows the Rayleigh test to be carried
out in its field of application (stability of the test between around 40
to 50 measurements after [-@bertran2002]).

@@

As proposed by McPherron (2018), a spatial investigation of the fabric
is proposed in a dedicated panel. Benn indices (1994) calculated for
each spatial series of remains are projected within a Benn diagram in
which each pole (isotropic, linear, and planar) is coloured respectively
in red, green and blue. The RGB colour code is reused to materialize the
fabric of the remains on spatial projections (plan and section views) of the material along the axes of the excavation grid. \bigbreak

@@

## Upload data

The simplest method to upload data in the application is to name the column names directly, as shown in the example files. If you miss a context column (*i.e.* 'id', 'code' or 'sample'), please create it by repeating a same character. By doing so, you do not need to select your variables in the main panel. The application correctly reads the data file directly.

@@

Five cases (*i.e.* forms of data) are covered by the application (fig.
\ref{fig:figure_upload_your_data}D):

@@

\begin{figure} [h]
\centering
  \includegraphics[width=16cm]{figure/upload_your_data.jpg}
  \caption{A. Browse your .csv file, B. choose your column separator, C. choose your decimal separator, D. check the right form of your dataset, E. to do so, have a look of example files, F. select the column needed in the application, G. a new dataset with new columns appears, your are ready to explore your data.}
\label{fig:figure_upload_your_data}
\end{figure}

@@

You can control the data you plot and the data appearing in the figures
and the summary data table by changing the checkboxes filters ('filter by sample'
and 'filter by code') and the slider (*i.e.* minimum number of measures) in the left
sidebar panel (fig. \ref{fig:figure_classical_method}A, B and C).
Certain geological processes can have a different impact on objects
depending on their physical properties (*e.g.* silex versus bones in a
water flow as in the example files provided where bones have a strong
preferential orientation and are much more 'linear' in the Benn diagram,
see [@thomas2019] and [@texier2020]). The natural processes models that
you can add to the Benn diagram are calculated using data mainly from
[@bertran1997; @bertran2002; @bertran2006; @lenoble2004]. You will find the reference model data (with corresponding litarature references) which is used in the application in the 'model reference data' subpanel. In the rose diagram, orientations are considered (and not directions) as the Rayleigh test is perform on orientations (Curray 1956) in the
summary table. You can download all the figures by selecting a file format (.pdf, .jpeg and .png) and pressing the 'download' button (fig. \ref{fig:figure_classical_method}E and F).

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/classical_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by codes, uncheck 'no code distinction', C. manage the minimum sample size to perform analyses, D. buld your model thanks to Bertran and Lenoble (2002) data, E. and F. download figures by selecting a file format and pressing the 'download' button, G. rose diagrams subpanel, H. Schmidt diagram subpanel, I. summary table with additional statistics}
\label{fig:figure_classical_method}
\end{figure}

@@

-   Benn diagrams with Benn indices computed with nearest neighbours points (McPherron 2018)
-   Spatial projections of the points according to their fabric shape
-   Summary data tables with orientation statistical tests, one at the
    scale of each spatial series (each object and its n nearest
    neighbours fabric = one row) and one at the scale of the sample (fig.
    \ref{fig:figure_spatialised_method}F and G).

@@

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/spatialised_method.jpg}
  \caption{A. filter by sample, B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics, D. make sticks in spatial projections, E. upload a georeferenced orthophoto and check 'include' orthophoto to display it, F. new dataset at the spatial series scale, G. summary table at the sample scale.}
\label{fig:figure_spatialised_method}
\end{figure}

@@

\begin{figure} [h]
\centering
  \includegraphics[width=17cm]{figure/spatial_exploration.jpg}
  \caption{A. if you want to filter by sample, uncheck 'No sample distinction' B. if you want to filter by code, uncheck 'no code distinction', C. choose the n nearest neighbors to perform the statistics (first Benn diagram), E. choose the n nearest neighbors to perform the statistics on the sampled data (second Benn diagram), F. make sticks in spatial projections, G. you can upload a georeferenced orthophoto and check 'include' orthophoto to display it, H. a new spatial exploration of the fabric is performed, I. new dataset at the selected spatial series scale, I. a similar spatial exploration is possible from the projections, K. new dataset at the selected spatial series scale.}
\label{fig:figure_spatial_exploration}
\end{figure}

@@

The 'spatialised' method generates often more linear fabrics in
comparison with the 'classical' method. This can be explained by the
fact that objects that are close in space are more likely to resemble
each other than objects that are further in space.

@@

For the reasons given above, we wish to create new models that
considered the particularities of the spatial method. A panel was
therefore set up to collect new data describing the fabric of
sedimentary processes in order to build a new "spatialised" reference
model (*cf.* the "classical" model proposed by Bertran and Lenoble
2002). Please, do not hesitate to send referential data to feed these new models ! \bigbreak

@@

The contributors panel aims at citing all contributors of this project
which is open to everyone interested in improving fabric analysis methods or
to contribute in building a new spatial model by sending reference data.
If any request or contribution, please contact the author of this
article. \bigbreak

-   use Fabryka $\beta$ 3.0:
    <https://marchaeologist.shinyapps.io/fabryka/> \smallbreak
-   source codes: <https://github.com/marchaeologist/fabryka/tree/main>
    \smallbreak
-   github reprository doi of Fabryka $\beta$ 3.0: \smallbreak

@@

The writing of the script was greatly facilitated by the work published
by Shannon McPherron (2018). It includes some functions published by the
author, with a few modifications. This work benefited from discussions
with Jean-Pierre Texier, Arnaud Lenoble, Pascal Bertran and Dominique
Todisco on issues related to site formation processes. The reference data used to build the models of the classical method were collected by P. Bertran throughout his career. P. Bertran provided us with all these data in order to feed the application. The developpement of the application also benefited from the help of Rémi Lemoy and François and Victor Baleux on mathematical issues. A first version of the application was tested by
Emmanuel Discamps and Aurélien Royer. I would like to thank all these reaseachers for their advice and
support. \bigbreak
